---
title: "2.crane_memory"
author: "Dennis Kim"
date: '2022-04-05'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages 
```{r packages, message =FALSE, warning = FALSE}
# data calling & wrangling 
library(here)
library(readr)
library(tidyr)
library(purrr)
library(dplyr)
library(forcats)
library(tibble)
library(stringr)
library(broom)

# movement analysis 
library(amt)

# visualization
library(ggplot2)
library(jtools)
library(ggstance)
library(broom.mixed)
```

# Load Data
```{r data call}
# read summer data for id == 5A
a515 = read_rds(here("data","summer_trk15_5A.Rdata"))
a516 = read_rds(here("data","summer_trk16_5A.Rdata"))
a517 = read_rds(here("data","summer_trk17_5A.Rdata"))
a518 = read_rds(here("data","summer_trk18_5A.Rdata"))
a519 = read_rds(here("data","summer_trk19_5A.Rdata"))
a520 = read_rds(here("data","summer_trk20_5A.Rdata"))
```

# Prepare data for SSF fitting
```{r ssf prep}
# factorize the age year
a515$age_yr <- factor(a515$age_yr, levels = c("hatch-yr", "second-yr", "third-yr", "fourth-yr", "fifth-yr", "sixth-yr", "adult"))
a516$age_yr <- factor(a516$age_yr, levels = c("hatch-yr", "second-yr", "third-yr", "fourth-yr", "fifth-yr", "sixth-yr", "adult"))
a517$age_yr <- factor(a517$age_yr, levels = c("hatch-yr", "second-yr", "third-yr", "fourth-yr", "fifth-yr", "sixth-yr", "adult"))
a518$age_yr <- factor(a518$age_yr, levels = c("hatch-yr", "second-yr", "third-yr", "fourth-yr", "fifth-yr", "sixth-yr", "adult"))
a519$age_yr <- factor(a519$age_yr, levels = c("hatch-yr", "second-yr", "third-yr", "fourth-yr", "fifth-yr", "sixth-yr", "adult"))
a520$age_yr <- factor(a520$age_yr, levels = c("hatch-yr", "second-yr", "third-yr", "fourth-yr", "fifth-yr", "sixth-yr", "adult"))

# drop the levels that are not present in the data
a515$nlcd_category <- droplevels(as.factor(a515$nlcd_category))
a516$nlcd_category <- droplevels(as.factor(a516$nlcd_category))
a517$nlcd_category <- droplevels(as.factor(a517$nlcd_category))
a518$nlcd_category <- droplevels(as.factor(a518$nlcd_category))
a519$nlcd_category <- droplevels(as.factor(a519$nlcd_category))
a520$nlcd_category <- droplevels(as.factor(a520$nlcd_category))

# change the reference level 
a515$nlcd_category <- fct_relevel(a515$nlcd_category, "wetlands") # wetlands 
a516$nlcd_category <- fct_relevel(a516$nlcd_category, "wetlands") # wetlands 
a517$nlcd_category <- fct_relevel(a517$nlcd_category, "wetlands") # wetlands 
a518$nlcd_category <- fct_relevel(a518$nlcd_category, "wetlands") # wetlands 
a519$nlcd_category <- fct_relevel(a519$nlcd_category, "wetlands") # wetlands 
a520$nlcd_category <- fct_relevel(a520$nlcd_category, "wetlands") # wetlands 

a515$age_yr <- fct_relevel(a515$age_yr, "hatch-yr")
a516$age_yr <- fct_relevel(a516$age_yr, "hatch-yr")
a517$age_yr <- fct_relevel(a517$age_yr, "hatch-yr")
a518$age_yr <- fct_relevel(a518$age_yr, "hatch-yr")
a519$age_yr <- fct_relevel(a519$age_yr, "hatch-yr")
a520$age_yr <- fct_relevel(a520$age_yr, "hatch-yr")

# calculate log sl and cos ta 
a515$log_sl_ <- log(a515$sl_)
a516$log_sl_ <- log(a516$sl_)
a517$log_sl_ <- log(a517$sl_)
a518$log_sl_ <- log(a518$sl_)
a519$log_sl_ <- log(a519$sl_)
a520$log_sl_ <- log(a520$sl_)

a515$cos_ta_ <- cos(a515$ta_)
a516$cos_ta_ <- cos(a516$ta_)
a517$cos_ta_ <- cos(a517$ta_)
a518$cos_ta_ <- cos(a518$ta_)
a519$cos_ta_ <- cos(a519$ta_)
a520$cos_ta_ <- cos(a520$ta_)
```


simplify the data 
```{r OD simplification}
# Set OD = 1 if the point is in a grid cell that was previously visited and 0 otherwise.
a516$OD15[a516$OD15 > 0] <- 1
a516$OD15[a516$OD15 < 0] <- 0
summary(a516$OD15)

a517$OD16[a517$OD16 > 0] <- 1
a517$OD16[a517$OD16 < 0] <- 0
summary(a517$OD16)

a518$OD17[a518$OD17 > 0] <- 1
a518$OD17[a518$OD17 < 0] <- 0
summary(a518$OD17)

a519$OD18[a519$OD18 > 0] <- 1
a519$OD18[a519$OD18 < 0] <- 0
summary(a519$OD18)

a520$OD19[a520$OD19 > 0] <- 1
a520$OD19[a520$OD19 < 0] <- 0
summary(a520$OD19)
```

# Further data investigation 
To ensure that 3 different measures of the locations and predictor in the data is reasonable
* OD predictor in each of the grid cells 
* counts of crane location in the same grid cells 
* counts of available locations (generated using amt) in each of the grid cells 

```{r visualization, fig.width=12,fig.height=10}
# filter out each data of OD and cases_
a516.df = a516 %>% select(OD15, case_, year) %>% rename(., ODprev = OD15)
a517.df = a517 %>% select(OD16, case_, year) %>% rename(., ODprev = OD16)
a518.df = a518 %>% select(OD17, case_, year) %>% rename(., ODprev = OD17)
a519.df = a519 %>% select(OD18, case_, year) %>% rename(., ODprev = OD18)
a520.df = a520 %>% select(OD19, case_, year) %>% rename(., ODprev = OD19)

# bind all the filtered data 
a5year = rbind(a516.df, a517.df, a518.df, a519.df, a520.df)

# save the data 
#write_rds(a5year, path = here("data", "summer_trk15_5A_avil_used_pts.Rdata"))

# visualization the proportion based on each year on the OD prev 
## count
ggplot(a5year, aes(x = as.factor(ODprev), group = case_))+
  geom_bar(position = position_dodge(), aes(fill = case_), stat = "count")+
  scale_fill_brewer(palette = "Paired")+
  labs(y = "Count", fill = "Used", x = "OD of previous year")+
  facet_wrap(~year, scales = "free")+
  ggtitle("(c) Distribution of used and available points in previous year's OD")

## proportion
ggplot(a5year, aes(x = as.factor(ODprev),group = case_))+
    geom_bar(position= position_dodge(), aes(y=..prop.., fill = case_), stat = "count")+
    scale_fill_brewer(palette="Paired")+
    geom_text(aes(label = scales::percent(round(..prop.., 2)),
                 y= ..prop.. ), stat= "count", vjust = -.3, position=position_dodge(0.9), size = 3) +
    labs(y = "Proportion", fill = "used", x = "OD of previous year")+
    facet_wrap(~year, scales = "free")
```

# Fitting SSF

**[note]** 

* **occurrence()** returns a distribution of where the animal might have been during the sampling period (or some range of times).
* **Negative coefficients** suggest cranes do not prefer using past year's locations.  
* **Positive coefficients** suggest they like to revisit past locations.

## Without landcover classes 
```{r no LC}
# same coefficient range compared to the model fits with LC 
#summary(fit_issf(case_ ~ OD15 + log_sl_ + strata(step_id_), data = a516)) # negative 
#summary(fit_issf(case_ ~ OD16 + log_sl_ + strata(step_id_), data = a517)) # positive 
#summary(fit_issf(case_ ~ OD17 + log_sl_ + strata(step_id_), data = a518)) # positive 
#summary(fit_issf(case_ ~ OD18 + log_sl_ + strata(step_id_), data = a519)) # positive 
#summary(fit_issf(case_ ~ OD19 + log_sl_ + strata(step_id_), data = a520)) # negative 
```


## 2016
```{r fit16}
# MODEL16: year 2016
m16 <- fit_issf(case_ ~ nlcd_category + OD15 + log_sl_ + strata(step_id_), data = a516)

summary(m16)
```
**[Interpreation]**

* OD15: **negative coefficient: In 2016, the crane does not select the locations where they were in last year 2015.**

## 2017
```{r fit17}
# MODEL17: year 2017
m17 <- fit_issf(case_ ~ nlcd_category + OD16 + log_sl_ + strata(step_id_), data = a517)

summary(m17)
```
**[Interpreation]**

* OD16: **positive coefficient: In 2017, the crane selects the locations where they were in last year 2016.**

## 2018
```{r fit18}
# MODEL18: year 2018
m18 <- fit_issf(case_ ~ nlcd_category + OD17 + log_sl_ + strata(step_id_), data = a518)

summary(m18)
```
**[Interpreation]**

* OD17: **positive coefficient: in 2018, the crane selects the locations where they were in last year 2017.**

## 2019
```{r fit19}
# MODEL19: year 2019
m19 <- fit_issf(case_ ~ nlcd_category + OD18 + log_sl_ + strata(step_id_), data = a519)

summary(m19)
```
**[Interpreation]**

* OD18: **positive coefficient: In 2019, the crane selects the locations where they were in last year 2018.**

## 2020
```{r fit20}
# MODEL20: year 2020
m20 <- fit_issf(case_ ~ nlcd_category + OD19 + log_sl_ + strata(step_id_), data = a520)

summary(m20)
```
**[Interpreation]**

* OD19: **negative coefficient: In 2020, the crane does not the locations where they were in last year 2019.**

# Summary 

Overall, The crane 5A would like to select their hatched (2015) locations during their 2nd year (2016) when they were by themselves without any information about the landscape. During their 3rd year (2017), the crane started exploring possible habitats to select and did not select the previous year's location (2016). After getting information about the landscape, the crane continue to revisit the same locations throughout the years (2018, 2019, 2020).  

# Visualization

## tidy the data 
First tidy the data and extract estimates and CIs of each model
```{r tidy the vis data}
# 2016
coef.m16 <- m16$model$coefficients %>% tidy() %>% mutate(year = "2016")
colnames(coef.m16) <- c("beta", "coefficient", "year")

## extract CIs for the model fit and create a new column of the beta 
conf.m16 <- m16$model %>% confint() %>% as.data.frame() %>% mutate(beta = c("nlcd_categorybarren","nlcd_categorycultivated", "nlcd_categorydeveloped","nlcd_categoryforest", "nlcd_categoryherbaceous","nlcd_categoryshrubland","nlcd_categorywater","OD15","log_sl_"))

## reorder by column name for left join
coef.m16 <- coef.m16[, c("year", "coefficient", "beta")]

## left join the confint df to the coefficient df 
fit.m16 <- left_join(coef.m16, conf.m16)
fit.m16


# 2017
coef.m17 <- m17$model$coefficients %>% tidy() %>% mutate(year = "2017")
colnames(coef.m17) <- c("beta", "coefficient", "year")

## extract CIs for the model fit and create a new column of the beta 
conf.m17 <- m17$model %>% confint() %>% as.data.frame() %>% mutate(beta = c("nlcd_categorycultivated", "nlcd_categorydeveloped","nlcd_categoryforest", "nlcd_categoryherbaceous","nlcd_categoryshrubland","nlcd_categorywater","OD16","log_sl_"))

## reorder by column name for left join
coef.m17 <- coef.m17[, c("year", "coefficient", "beta")]

## left join the confint df to the coefficient df 
fit.m17 <- left_join(coef.m17, conf.m17)
fit.m17


# 2018
coef.m18 <- m18$model$coefficients %>% tidy() %>% mutate(year = "2018")
colnames(coef.m18) <- c("beta", "coefficient", "year")

## extract CIs for the model fit and create a new column of the beta 
conf.m18 <- m18$model %>% confint() %>% as.data.frame() %>% mutate(beta = c("nlcd_categorycultivated", "nlcd_categorydeveloped","nlcd_categoryforest", "nlcd_categoryherbaceous","nlcd_categoryshrubland","nlcd_categorywater","OD17","log_sl_"))

## reorder by column name for left join
coef.m18 <- coef.m18[, c("year", "coefficient", "beta")]

## left join the confint df to the coefficient df 
fit.m18 <- left_join(coef.m18, conf.m18)
fit.m18

# 2019
coef.m19 <- m19$model$coefficients %>% tidy() %>% mutate(year = "2019")
colnames(coef.m19) <- c("beta", "coefficient", "year")
coef.m19

## extract CIs for the model fit and create a new column of the beta 
conf.m19 <- m19$model %>% confint() %>% as.data.frame() %>% mutate(beta = c("nlcd_categorycultivated", "nlcd_categorydeveloped","nlcd_categoryforest", "nlcd_categoryherbaceous","nlcd_categoryshrubland","nlcd_categorywater","OD18","log_sl_"))

## reorder by column name for left join
coef.m19 <- coef.m19[, c("year", "coefficient", "beta")]

## left join the confint df to the coefficient df 
fit.m19 <- left_join(coef.m19, conf.m19)
fit.m19

# 2020
coef.m20 <- m20$model$coefficients %>% tidy() %>% mutate(year = "2020")
colnames(coef.m20) <- c("beta", "coefficient", "year")
coef.m20

## extract CIs for the model fit and create a new column of the beta 
conf.m20 <- m20$model %>% confint() %>% as.data.frame() %>% mutate(beta = c("nlcd_categorycultivated", "nlcd_categorydeveloped","nlcd_categoryforest", "nlcd_categoryherbaceous","nlcd_categoryshrubland","nlcd_categorywater","OD19","log_sl_"))

## reorder by column name for left join
coef.m20 <- coef.m20[, c("year", "coefficient", "beta")]

## left join the confint df to the coefficient df 
fit.m20 <- left_join(coef.m20, conf.m20)
fit.m20


# bind all the data
fit.M <- rbind(fit.m16, fit.m17, fit.m18, fit.m19, fit.m20)
fit.M

## change the column names 
colnames(fit.M) <- c("year", "coefficient", "beta", "lowCI", "highCI")

# reorder by column name 
fit.M <- fit.M[, c("beta", "year", "coefficient", "lowCI", "highCI")]

# change the row name for beta 
fit.M$beta <- gsub("nlcd_category","", fit.M$beta) 
fit.M$beta <- gsub("OD15","ODprev", fit.M$beta) 
fit.M$beta <- gsub("OD16","ODprev", fit.M$beta)
fit.M$beta <- gsub("OD17","ODprev", fit.M$beta)
fit.M$beta <- gsub("OD18","ODprev", fit.M$beta)
fit.M$beta <- gsub("OD19","ODprev", fit.M$beta)
fit.M$beta <- gsub("log_sl_","logSL", fit.M$beta)

fit.M
```

```{r visualization}
# plot the coefficient visualization
## drop the rows with barren and herbaceous (their estimates are bigger range than the other predictors) - hard to see the trends for the other predictors 
fit.M1 <- subset(fit.M, beta != "barren" & beta != "herbaceous")

## save the estimates df 
#write_rds(fit.M1, path = here("data", "5A_summer_estimates.Rdata"))

# re-order the beta labels 
coef.vis <- fit.M1 %>% dplyr::mutate(beta = forcats::fct_relevel(beta, sort)) %>% 
ggplot(., aes(coefficient, beta, colour = beta)) +
  geom_vline(xintercept = 0, lty=2, lwd=1, colour = "grey50")+
  geom_point(size = 3, pch = 20, position = position_dodge(width = 0.5))+
  geom_linerange(aes(xmax = highCI, xmin = lowCI))+
  xlim(-5, 5)+
  theme_bw()+
  facet_grid(~ year)+
  xlab("Effect size")+
  theme(legend.position = "none",
        axis.title.y = element_blank())

coef.vis
```

# Document Footer 
```{r session info}
sessionInfo()
```