---
title: "Mdeer_Merkle_2019"
author: "Dennis Kim"
date: "2023-01-17"
output: html_document
---

# Objective 

Script to demonstrate how to set up SSF from Merkle et al. 2019 (Ecol Lett)
Written by Jerod Merkle, January 2023 

# Document Preamble 
```{r preamble, include = FALSE}
# load libraries
require(tidyverse)
require(sf)
require(raster)
require(lubridate)
library(amt)
library(knitr)
library(here)

options(width = 150)

# set knitr options 
opts_chunk$set(fig.width = 6, fig.height = 5, comment = NA)
```

# Data Preparation

## Positional data

Read in positional data - this is 1 individual mule deer collared for nearly 2 years in NW Wyoming. These data were collected by Hall Sawyer (WEST Inc)
```{r position data}
dat_all <- readRDS(here::here("data", "GPSdata_muldeer445.rds"))
```

## Movement data 

import or develop table of migration start and stop dates for each year and season
```{r movement data}
migtable <- data.frame(id=445, 
                       year=c(2014, 2014, 2015, 2015),
                       season=c("spring","fall","spring","fall"),
                       date_start=c("2014-05-04","2014-11-02","2015-04-22","2015-11-13"),
                       date_end=c("2014-05-17","2014-11-21","2015-05-13","2015-11-24")) %>% 
  mutate(date_start=ymd(date_start, tz="America/Denver"),
         date_end=ymd(date_end, tz="America/Denver"))

# add column to denote year of monitoring
migtable <- migtable %>% 
  group_by(id) %>% 
  arrange(year, desc(season)) %>% 
  mutate(year_cumm=year-min(year)+1)
```


Isolate data to be analyzed (i.e., analyze second year of data, while using the first year to index of memory)
Here, we grab the spring migration for the second year of monitoring
```{r data isolation}
dat2keep <- migtable %>% 
  filter(year_cumm==2, season=="spring")

dat <- do.call(rbind, lapply(1:nrow(dat2keep), function(i){
  dat_all %>% 
    filter(id==dat2keep$id[i], 
           date >= dat2keep$date_start[i],
           date <= dat2keep$date_end[i]) %>% 
    mutate(season="spring") %>% 
    return()
}))
```

## Prep indices of memory 

*1) Isolate previous year's migration routes*

Loop through each id/season identified in dat2keep
```{r previous migration routes}
prev.routes <- do.call(rbind, lapply(1:nrow(dat2keep), function(i){
  
  # get migtable organized for the previous year of data but for the id in question
  mig.i <- migtable %>% 
    filter(id==dat2keep$id[i], year_cumm==1, season=="spring")
  
  # isolate the points for the previous year's migration in the same season
  pts <- dat_all %>% 
    filter(id==mig.i$id, 
           date >= mig.i$date_start,
           date <= mig.i$date_end)
  
  # turn into a sf line dataframe
  ln <- pts %>% 
    st_combine() %>% 
    st_cast("LINESTRING") %>% 
    data.frame(id=mig.i$id, 
               season="spring",
               year=mig.i$year,
               geometry=.) %>% 
    st_as_sf(sf_column_name = "geometry")
  
  return(ln)
}))
```

*2) Isolate centroid of previous year's summer range*

Again, loop through each id/season identified in dat2keep
```{r previous summer range}
prev.ranges <- do.call(rbind, lapply(1:nrow(dat2keep), function(i){
  
  # get migtable organized for the previous year of data but for the id in question
  mig.i <- migtable %>% 
    filter(id==dat2keep$id[i], year_cumm==1)
  
  # isolate the points for the period bewteen the end of the spring migration
  # and the start of the fall migration (i.e., summer)
  pts <- dat_all %>% 
    filter(id==mig.i$id, 
           date >= mig.i$date_end[mig.i$season=="spring"],  
           date <= mig.i$date_start[mig.i$season=="fall"])
  
  # calculate centroid and turn into a sf point dataframe
  centr <- pts %>% 
    st_union() %>% 
    st_centroid() %>% 
    data.frame(id=mig.i$id[1], 
               season="summer",
               year=mig.i$year[1],
               geometry=.) %>% 
    st_as_sf(sf_column_name = "geometry")
  
  return(centr)
}))

# remove the data that are no longer needed for further analysis
rm(dat2keep, migtable)
```

# AMT data conversion 

## Track conversion by amt package 
Translate location data to tracks - Note: from here on out, there is no code to handle multiple individuals 
```{r amt transition}
# make tracks
trk <- dat %>% 
  mutate(x = st_coordinates(.)[,1], # add x and y as columns
         y = st_coordinates(.)[,2]) %>% 
  st_drop_geometry() %>%   #need to remove the geometry column
  make_track(x, y, date, id = id, season = season, 
             crs = st_crs(dat))

# check the tracks
trk %>% summary()

# transform into a crs with meters as unit
trk <- trk %>% 
  transform_coords(5072)
```

## Step lengths 
Create step lengths (both used and random)
```{r steps}
# have a look at step lengths
trk <- trk %>% 
  mutate(sl_ = step_lengths(.))

hist(trk$sl_)
median(trk$sl_, na.rm=TRUE)

# sampling rate is very consistent so don't need to worry about resampling  
summarize_sampling_rate(trk) 

# create steps
ssf1 <- trk %>% 
  steps()

ssf1 %>% head()

# check to see if some directional persistence is present
hist(ssf1$ta_) # indeed there is, so the SSF is properly set up!

# draw random steps
ssf1 <- ssf1 %>% 
  random_steps()

ssf1 %>% head()
```

# Extract habitat covariates 

```{r add habitat covariates}
# read in variables first
# slope derived from USGS NED at approx 30m resolution
slope <- raster(here::here("data", "Slope30m.tif"))

# Terrain Position Index derived from USGS NED at approx 30m resolution
tpi <- raster(here::here("data", "TPI_9x9window_30m.tif"))

# TRASP derived from USGS NED at approx 30m resolution calculated based on
# trasp = ( 1 - cos((pi/180)(a-30) ) / 2 where; a = aspect in degrees
# values near 1 are SSW slopes and values near 0 are NNE. 
# Note: to save space, values were multiplied by 100 and saved as integers
trasp <- raster(here::here("data", "TRASP30m.tif"))

# Tree cover from the National Landcover Dataset (2019): https://www.mrlc.gov/
treecov <- raster(here::here("data", "TreeCover_nlcd2019_30m.tif"))

# Biomass of Forbs and grasses from the RangeLand Analysis Platform
# year 2015, units are lbs/acre (https://rangelands.app/)
herbs <- raster(here::here("data", "RAP_2015_Biomass_ForbsGrasses.tif"))

# Need to build separate sf object for extraction because amt cannot deal with different projections of the environmental data
step_ends <- ssf1 %>% 
  as.data.frame() %>% 
  select(x2_, y2_) %>% 
  st_as_sf(coords=c("x2_","y2_"), dim="XY", crs=5072)

# habitat extraction
ssf1$tpi <- raster::extract(tpi, st_transform(step_ends, projection(tpi)))
ssf1$trasp <- raster::extract(trasp, st_transform(step_ends, projection(trasp)))
ssf1$slope <- raster::extract(slope, st_transform(step_ends, projection(slope)))
ssf1$treecov <- raster::extract(treecov, st_transform(step_ends, projection(treecov)))
ssf1$herbs <- raster::extract(herbs, st_transform(step_ends, projection(herbs)))

# remove the data that are no longer needed for further analysis
rm(tpi, trasp, slope, treecov, herbs)
```

# Extract memory components 

## 1) distanct to previous route 
```{r dis to prev route}
prev.routes <- prev.routes %>% 
  st_transform(5072)  # transform to projection of the tracks

# calculate the distance between step ends and previous routes
ssf1$dist2prevroute <- as.numeric(st_distance(step_ends, prev.routes))
```

## 2) direction to previous summer range 
```{r dir to prev summer range}
prev.ranges <- prev.ranges %>% 
  st_transform(5072) %>% # transform to projection of the tracks
  mutate(x = st_coordinates(.)[,1], # add x and y as columns
         y = st_coordinates(.)[,2])

# calculate direction from source location to previous range (in radians)
direct2prev_rng_source <- atan2((prev.ranges$y-ssf1$y1_), (prev.ranges$x-ssf1$x1_))

# calculate direction from source location to target location (in radians)
direct2target <- atan2((ssf1$y2_-ssf1$y1_), (ssf1$x2_-ssf1$x1_)) # change this to direct2target

# subtract the two and take the cosine
ssf1$direct2prev_rng <- cos(direct2prev_rng_source-direct2target)
# Values of 1 represent a step directly towards previous range
# Values of 0 represent a step directly away from previous range
```

# Fit SSFs 

## 1) Only environmental model
```{r env ssf model}
# add log of step length
ssf1 <- ssf1 %>% 
  mutate(sl_log = log(sl_))

# fit environmental model
m_envir <- ssf1 %>% 
  fit_clogit(case_ ~ sl_log+tpi+slope+trasp+herbs+treecov+
               strata(step_id_))

summary(m_envir)
```
**Interpretation:** This deer select steps that go to more ridgy areas, warmer aspects with higher biomass of herbs

## 2) Memory + Environmental model
```{r env mem ssf model}
# fit environmental model + memory
m_envir_memor <- ssf1 %>% 
  fit_clogit(case_ ~ sl_log+tpi+slope+trasp+herbs+treecov+
               dist2prevroute+direct2prev_rng+
               strata(step_id_))

summary(m_envir_memor)
```
**Interpretation:** This deer select steps that go to more ridgy areas, warmer aspects with higher biomass of herbs and they select steps closer to their previous route and in the direction of their previous summer range

# compare relative empirical support for each model
```{r AIC}
AIC(m_envir)
AIC(m_envir_memor) 
```

# Footer
```{r footer}
sessionInfo()
```