---
title: "1.amt_conv"
author: "Dennis Kim"
date: "2022-10-06"
output: html_document
---

# Objective 

To explore the structure of the bear data set. To do this, I will: 

1. Generate used and available locations of the bear through amt framework. 
2. Extract habitat covariate from habitat layer. 
3. Create sf polygons that define the buffer of each grid 
4. Calculate the points that fall within each buffer and save it as each raster layer 
5. Calculate the TSLV similar to Uli's approach 
6. Include TSLV values per each location and save them in the column of the data 
7. Fit SSF model with a TSLV predictor 

## Document Preamble 
```{r preamble, include = FALSE}
# load libraries
library(knitr)
library(dplyr)
library(readr)
library(data.table)
library(DT)
library(here)
library(stringr)
library(tidyr)
library(purrr)
library(amt)
library(ggplot2)
library(raster)
library(sf)
library(recurse)
library(scales)
library(sp)
library(geosphere)
library(generics)
library(sfheaders)
library(lubridate)
options(width = 150)

# set knitr options 
opts_chunk$set(fig.width = 6, fig.height = 5, comment = NA)
```

## Tracking data 

Read in the gps data 
```{r read gps data}
# Location data of the bear 
bear <- read.csv(here::here("data", "GF1143_raw.csv"))

# change the time format of the data 
bear$datetime <- as.POSIXct(bear$datetime, format =  "%Y-%m-%d %H:%M")

# summary of the data
summary(bear)
```

Plot the data 
```{r bear location vis}
ggplot(bear, aes(x = x, y = y))+ 
  geom_point()
```

## Prepare environmental data 

Add environmental covariate (berry)
```{r habitat layer}
# call the berries layer
berry <- raster(here("data/berries/berries.tif"))

# plot the layer
plot(berry)

# crs of the layer 
crs(berry)
```


## amt conversion 

Add a track class to the data and summarize the data 
```{r amt conversion}
# make tracks 
trk.bear <- amt::make_track(bear, .x=x, .y=y, .t=datetime, crs=CRS("+init=epsg:26908"))

summary(trk.bear)
```

Summarize the sampling rates of the bear
```{r sampling rates}
# 4 hour sampling seems reasonable - no need to resample - since it is already resampled 
summarize_sampling_rate(trk.bear) 
```

change the track point to step 
1. Resample track and filter bursts 
2. Convert track to steps 
3. Create random steps 
4. Extract covariate values 
```{r retrk bear, echo = FALSE}
# follow the above approach 
ssfdat.bear <- 
  trk.bear %>% track_resample(rate = hours(4), tolerance = minutes(30)) %>% 
  steps_by_burst() %>% 
  random_steps(n_control = 1) %>% 
  extract_covariates(berry) %>% 
  filter(!is.na(ta_))

# summary of the ssf data
summary(ssfdat.bear)
```

Check the reasonable buffer distance for each step length - I will pick 500m as a buffer distance from the step lengths 
```{r sl distribution}
ssfdat.bear %>% filter(case_ == "TRUE") %>% dplyr::select(sl_) %>% summary()
```

## Time Since Last Visit 

### Create buffer zones of the step lengths

select randomized and used locations of the tracks from the ssfdat.bear
```{r st points, echo = FALSE}
# select random locations that matched with observed step lengths 
obs <- ssfdat.bear %>% filter(case_ == "TRUE")
obs

ran <- ssfdat.bear %>% filter(case_ == FALSE)
matched.ran <- subset(ran, x1_ %in% obs$x1_)
matched.ran


# select observed locations
obs.loc <- obs %>% dplyr::select(x1_, y1_)
colnames(obs.loc) <- c("x", "y")
obs.loc.sf <- st_as_sf(obs.loc, coords = c("x", "y"), crs = "+proj=utm +zone=8 +datum=NAD83 +units=m +no_defs")
obs.loc.sf %>% head()

# select random locations
ran.loc <- matched.ran %>% dplyr::select(x2_, y2_)
colnames(ran.loc) <- c("x", "y")
ran.loc.sf <- st_as_sf(ran.loc, coords = c("x", "y"), crs = "+proj=utm +zone=8 +datum=NAD83 +units=m +no_defs")
ran.loc.sf %>% head()
```

create a sf linestring object representing a step length between 2 observed locations within the whole trajectory 
```{r time since last visit calc, echo = FALSE}
# select observed locations only    
obs.step <- obs %>% dplyr::select(x1_,y1_, x2_, y2_)
colnames(obs.step) <- c("x1", "y1", "x2", "y2")
head(obs.step)

# make a sf linestring representing each step length
## functions for making sf linestrings
make_line <- function(xy2){
    st_linestring(matrix(xy2, nrow=2, byrow=TRUE))
}

make_lines <- function(df, names=c("x1","y1","x2","y2")){
    m = as.matrix(df[,names])
    lines = apply(m, 1, make_line, simplify=FALSE)
    st_sfc(lines)
}

sf_pts_to_lines <- function(df, names=c("x1","y1","x2","y2")){
    geom = make_lines(df, names)
    df = st_sf(df, geometry=geom)
    df
}

# make linestrings (step length) geometry
sl_df <- sf_pts_to_lines(obs.step) %>% dplyr::select(geometry)

# set the crs 
st_crs(sl_df) <- "+proj=utm +zone=8 +datum=NAD83 +units=m +no_defs"

head(sl_df)

# plot the SL geometry
plot(sl_df)
```

calculate the buffer area with the 500m distance from each focal step length 
```{r sl intersection, echo = FALSE}
# create 500 m buffer zone of each step length
sl.buffer = st_buffer(sl_df, 500)

# plot the sl buffer 
plot(sl.buffer)
```

Follow Uli's approach: Their definition of TSLV is short and sweet (see equation 4 in that paper) - basically, it's 0 if the point in question is within some distance δ (this value could be similar to the value you used for your buffer) of the previous point, and otherwise it's (previous TSLV + 1). So they define it iteratively, starting at the first point and iteratively updating TSLV with each time step.

### Observed locations with updated TSLV

Observed points within the step length buffers
```{r obs: yes within or no, echo = FALSE}
# check if it is working properly by plotting the first sl buffer with points
plot(sl.buffer$geometry[1]) # select the buffer of the first step length
plot(obs.loc.sf$geometry, add = TRUE) # put all the locations to see if some of them are within the location

# identify if any points are within each step length bufferzone 
obs.within.buffer <- st_contains(sl.buffer, obs.loc.sf)
# convert to df 
obs.within.buffer <- obs.within.buffer %>% as.data.frame()
# change the colnames - the inital df will have 2 columns called row.id and col.id - this represnts row.id as buffer id and col.id as pt id that is counted within the buffer 
colnames(obs.within.buffer) <- c("buffer.id", "pt.id")
# check the data
head(obs.within.buffer)
```
Initial prep for the tslv calculation
```{r obs: tslv prep, echo = FALSE}
# nest by each unique point 
nest.obs.pts <- obs.within.buffer %>% nest_by(pt.id)

# last visit time step 
nest.obs.pts$last.visit.time.step = c(0:533)

# nest by each unique buffer
nest.obs.buf <- nest.obs.pts  %>% unnest(data) %>% ungroup(pt.id) %>% nest_by(buffer.id)

# current visit time step
nest.obs.buf$current.time.step = c(1:534)

# unnest the data and call it as comp.buffer
comp.obs.buffer <- nest.obs.buf %>%  unnest(data) %>% ungroup(buffer.id)
comp.obs.buffer %>% head()
```

Create a spatial-temporal cognitive map that track the tslv value per point
```{r obs: calculate updated tslv, echo = FALSE}
# group by pt.id and apply the functions we created to account the updated tslv values and nest by pt.id
obs.tslv <- comp.obs.buffer %>% group_by(pt.id) %>% 
  mutate(tslv = 
           ifelse(
             # if buffer id falls into less 101, then go to case 1 - no? then go to case 2 
             buffer.id < 101,
             
             # case 1: if current time step is equal or greater than the last visit time step by 1, then set it as 0
             ifelse(current.time.step == last.visit.time.step | current.time.step == (last.visit.time.step+1), 0, 
             
             # otherwise, calculate the time between the last visit to the current time step 
             current.time.step - last.visit.time.step), 
             
             # case 2: if current time step is equal or greater than the last visit time step by 1, then set it as 0
             ifelse(current.time.step == last.visit.time.step | current.time.step == (last.visit.time.step+1), 0, 
             # otherwise, set the first last visit time step as 100 
             ifelse(row_number(last.visit.time.step) == 1, 100, 
             # otherwise, calculate the time between the last visit to the current time step            
             current.time.step - (last.visit.time.step)
             ))))%>% 
  # create an actual updated TSLV by subtracting the earlier tslv value 
  mutate(updated.tslv = tslv - lag(tslv)) %>%
  # only select the latest updated tslv value per id
  slice_tail(n = 1)

# NA values indicate that there is no updated tslv values
obs.tslv <- obs.tslv %>% mutate(updated.tslv = coalesce(updated.tslv, tslv))

# summary(obs.tslv)
obs.tslv %>% summary()

obs.tslv
```

### random locations with updated TSLV

random points within the step length buffers
```{r random: yes within or no, echo = FALSE}
# check if it is working properly by plotting the first sl buffer with points
plot(sl.buffer$geometry[1]) # select the buffer of the first step length
plot(ran.loc.sf$geometry, add = TRUE) # put all the locations to see if some of them are within the location

# identify if any points are within each step length bufferzone 
ran.within.buffer <- st_contains(sl.buffer, ran.loc.sf)
# convert to df 
ran.within.buffer <- ran.within.buffer %>% as.data.frame()
# change the colnames - the inital df will have 2 columns called row.id and col.id - this represnts row.id as buffer id and col.id as pt id that is counted within the buffer 
colnames(ran.within.buffer) <- c("buffer.id", "pt.id")
# check the data
head(ran.within.buffer)
```
Initial prep for the tslv calculation
```{r randOm: tslv prep, echo = FALSE}
# nest by each unique point 
nest.ran.pts <- ran.within.buffer %>% nest_by(pt.id)

# last visit time step 
nest.ran.pts <- nest.ran.pts %>% unnest(data) %>% mutate(last.visit.time.step = dplyr::lag(pt.id, n = 1, deafault = NA)) 

# replace na in last.visit.time.step
nest.ran.pts$last.visit.time.step[which(nest.ran.pts$last.visit.time.step %>% is.na())] <- 0

# nest by each unique buffer
nest.ran.buf <- nest.ran.pts %>% ungroup(pt.id) %>% nest_by(buffer.id)

# current visit time step (NOTE:: make sure to update this!!! based on how many random steps it generates!)
nest.ran.buf$current.time.step = c(1:513)

# unnest the data and call it as comp.buffer
comp.ran.buffer <- nest.ran.buf %>%  unnest(data) %>% ungroup(buffer.id)
comp.ran.buffer %>% head()
```

Create a spatial-temporal cognitive map that track the tslv value per point
```{r random: calculate updated tslv, echo = FALSE}
# group by pt.id and apply the functions we created to account the updated tslv values and nest by pt.id
ran.tslv <- comp.ran.buffer %>% group_by(pt.id) %>% 
  mutate(tslv = 
           ifelse(
             # if buffer id falls into less 101, then go to case 1 - no? then go to case 2 
             buffer.id < 101,
             
             # case 1: if current time step is equal or greater than the last visit time step by 1, then set it as 0
             ifelse(current.time.step == last.visit.time.step | current.time.step == (last.visit.time.step+1), 0, 
             
             # otherwise, calculate the time between the last visit to the current time step 
             last.visit.time.step - current.time.step), 
             
             # case 2: if current time step is equal or greater than the last visit time step by 1, then set it as 0
             ifelse(current.time.step == last.visit.time.step | current.time.step == (last.visit.time.step+1), 0, 
             # otherwise, set the first last visit time step as 100 
             ifelse(row_number(last.visit.time.step) == 1, 100, 
             # otherwise, calculate the time between the last visit to the current time step            
             last.visit.time.step - current.time.step
             ))))%>% 
  # create an actual updated TSLV by subtracting the earlier tslv value 
  mutate(updated.tslv = tslv - lag(tslv)) %>%
  # only select the latest updated tslv value per id
  slice_tail(n = 1)

# NA values indicate that there is no updated tslv values
ran.tslv <- ran.tslv %>% mutate(updated.tslv = coalesce(updated.tslv, tslv))

# check the data
summary(ran.tslv)
```

### Merge the data

Merge the updated tslv information to each observed and random point data
```{r merge tslv, echo = FALSE}
# Observed

# create an id column for the points 
obs.loc.sf <- obs.loc.sf %>% mutate(pt.id = c(1:534))

# only select the applied ids 
obs.tslv.df <- obs.tslv %>% dplyr::select(pt.id, updated.tslv)

# Random - Apply same below 
ran.loc.sf <- ran.loc.sf %>% mutate(pt.id = c(1:534))

ran.tslv.df <- ran.tslv %>% dplyr::select(pt.id, updated.tslv)
```

merge the tslv information to the point data 
```{r convert sf to df tslv, echo = FALSE}
# observe 
obs.df <- left_join(obs.loc.sf, obs.tslv.df)

# random
ran.df <- left_join(ran.loc.sf, ran.tslv.df)

# convert geometry back to lat and long
## observed
obs.df1 <- obs.df %>% mutate(x = unlist(map(obs.df$geometry,1)),
                   y = unlist(map(obs.df$geometry,2))) %>% dplyr::select(x, y, updated.tslv) %>% st_drop_geometry()

## random
ran.df1 <- ran.df %>% mutate(x = unlist(map(ran.df$geometry,1)),
                   y = unlist(map(ran.df$geometry,2))) %>% dplyr::select(x, y, updated.tslv) %>% st_drop_geometry() 
```

include tslv information to the original data
```{r amt tslv, echo = FALSE}
# extract updated tslv values per observed and random locations 
obs.loc.TSLV <- left_join(obs.loc, obs.df1) %>% mutate(case_ = TRUE)
colnames(obs.loc.TSLV) <- c("x1_", "y1_", "tslv", "case_")
obs.loc.TSLV %>% head()

ran.loc.TSLV <- left_join(ran.loc, ran.df1) %>% mutate(case_ = FALSE)
colnames(ran.loc.TSLV) <- c("x2_", "y2_", "tslv", "case_")
ran.loc.TSLV %>% head()

# merge the updated TSLV values per observed and random locations in ssf data 
## first merge the observed locations TSLV
ssfdat.bear1 <- left_join(ssfdat.bear, obs.loc.TSLV)
ssfdat.bear1 %>% head()

## second merge with the random locations tslv
ssfdat.bear2 <- left_join(ssfdat.bear1, ran.loc.TSLV, by = c("x2_", "y2_", "case_"))
ssfdat.bear2 %>% head()

# unite the tslv columns - replace na to 0 since there is no tslv for the locations
ssfdat.bear.final <- ssfdat.bear2 %>% mutate(TSLV = coalesce(tslv.x, tslv.y)) %>% replace_na(list(TSLV = 0)) %>% dplyr::select(-c(tslv.x, tslv.y))
ssfdat.bear.final %>% head()

# save the data 
#write_rds(ssfdat.bear.final, path = here("data", "ssf_dat.Rdata"))
```


## FitSSF
```{r fit ssf}
# only resource predictor 
ssfdat.bear.final %>% amt::fit_issf(case_ ~ berries+ log(sl_)+ cos(ta_)+ strata(step_id_)) %>% summary()

# TSLV predictor included
ssfdat.bear.final %>% amt::fit_issf(case_ ~ berries+ TSLV+ log(sl_)+ cos(ta_)+ strata(step_id_)) %>% summary()

# Interaction between TSLV and berries
ssfdat.bear.final %>% amt::fit_issf(case_ ~ berries+ TSLV+ TSLV:berries+ log(sl_)+ cos(ta_)+ strata(step_id_)) %>% summary()
```
*Interpretation*

- TSLV: the general idea is that animals may avoid recently used areas. This may be for many reasons - e.g., a predatory animal like a wolf will not hunt in the same patch because the prey animals there will get used to the presence of a predator and hide. Or alternatively, a grazing forager may exhaust all the vegetation in a patch and be forced to move elsewhere to eat. In areas where the animal has not been recently, it's more likely that resources will be "refreshed", making them more attractive to visit. And of course it is spatial (or spatio-temporal) memory that is driving the animal's ability to navigate to these locations. All in all, this is quite similar to what you have suggested but I figured it's good to provide biological background for what the hypotheses behind this model are.

- TSLV:berries: animals move more slowly in areas that they haven't visited in a while where the berries are present (negative interaction)

## Footer
```{r footer}
sessionInfo()
```